Learning how to act: making good decisions with machine learning

### 2017-05-19 thesis/thesis_map.tex
draft of chapter 2 complete

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib}
\usepackage{wrapfig}

@@

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\cf}[2]{{#1}_{#2}}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

@@

@@

\begin{document}
\todo{figure out theoremstle is undefined}
\def\ci{\perp\!\!\!\perp} % from Wikipedia

@@

3100 words at the start of boot camp.

\chapter*{Random notes}

\begin{figure}[h]
\centering
\caption{Causal inference without a framework}
\label{fig:mutilatednet}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{original network, $G$}
\label{fig:mutilatedOriginal}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};

\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\caption{network after an intervention $G^*$}
\label{fig:mutilatedAfter}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$x$};
\node[main node, below right=of 1](3){$Y$};

\begin{figure}[h]
\caption{}
\label{fig:causal_inference_without_do}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){Treatment};
\node[node, right=of 2](3){Outcome};
\node[lt, above=of 3](4){U};
\path[]
	(2) edge (3)
	(4) edge (3);

There is a connection here to the degree to which the model is invariant. A model that is invariant to more things is more explanatory. We could look at levels of hierarchy. There is causal graphical models, which are invariant to interventions on variables. Can I tie this to a physical system and give more in depth examples of the nested tier of hierarchies?

I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why.

Sometimes the causal component is obvious.

Relationship to generalisation. Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time. The assumption is there are less places for change to come in.

If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.

The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)? This may be a reason for using structure learning type algorithms even if you are simply doing prediction.

\subsubsection*{The role of assumptions in causal inference}

Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.

Does predictive accuracy indicate a good causal model?

Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.

Generalising the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parachutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalise). (and given sufficient generalisability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)

Is causal inference possible (are the conclusions ever valid)?

Only from randomised experiment ... if people can do it without recourse to constant randomised experiment then an algorithm exits.

The challenges and limitations of inferring causal effects from observation data are numerous. However, there are problems we need to solve. The questions is are these problems so serious as to warrant disregarding all non-experimental data. I think answer is a clear no. Need ongoing validation, with assumptions. We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).

\subsection*{Relationship to interpretability}

A desire for interpretability indicates that something has been left out of the loss function.

Are causal/interpretable models more reliable or more likely to generalise well?

One form of interpretability gives people insight into what the features are that the model is relying on.

If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.

Specifically, people can
\begin{itemize}
\item rule out many possible features as highly unlikely to be relevant to a problem
\end{itemize}

People have access to a lot of detailed prior knowledge.

\subsection*{Relationship to transfer learning}

find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time). Causal models predict the outcome of actions. We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike. So we have to make representations such that things are stable.

This is tightly related to generalisability. If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results. We don't include , the color of the doctors shirt, the size of the room, ...

For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales. However, this relationship may be very unstable over time because the ad slots are sold at auction. The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword. However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.

Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.

@@

\section{What is causality and why do we care? (2000 words)}

@@

\subsection{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (REFERENCES)
\item what is explanation?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}
\subsection{Overview of this thesis}
\subsection{Identifying when we have a causal problem}

@@

\section{Approaches to causality (1000 words)}

@@

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds. We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.  
\end{example}

@@

In example \ref{exm:adjusting} we are interested $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing pre-school $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonalble to assume that attending could pre-school effect the likelyhood of graduating from high school and that your parents socio-economic status effects both pre-school attendance and high-school graduation. If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal bayesian network in figure \ref{exm:adjusting}. In this case, the interventional distribution is not equivelent to the observational one. If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool. To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group, $\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}$.

@@

In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$. The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated. The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment. This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.

Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.  Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution we would see of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$. Now if $X=0$ then $Y = \cf{Y}{0}$. Equivalently stated:

\eq{
\P{Y^{0}|X=0}&= \P{Y|X=0}\\
\P{Y^{1}|X=1}&=P{Y|X=1}
}

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignore-ability assumptions \cite{Rosenbaum1983}. They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated. Again this is justified in our example due to the randomisation of treatment assignment. In general these assumption do not hold. If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model. that influenced both the decision to get treatment and the likelihood of recovery given treatment or placebo.

\todo{SUTVA assumption}

There are some complex philosophical objections to counter factuals arising from the way they describe alternate universes that were never realised and are not empirically testable \ref{XXXXX}. This can result in some concrete problems. Consider the following example based on  \ref{Dawid2000}.

Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$. Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal. We will assume equal variance for simplicity.

% vector with mean mu_t,mu_c and a covariance matrix
\eq{
P(Y_t,Y_c) \sim N()
}

then their difference is also normal,

\eq{
P(\tau) = N(\mu_t - \mu_c,2\sigma^{2}(1-\rho))
}

\includegraphics[scale=.6]{figures/counterfactual_nonidentify.pdf}

Key issue is that we can never observe the joint distribution over $Y_t$ and $Y_c$. As a result, the variance of $\tau$ is not identifiable, even from experimental data. It seems on the face of it that $\tau$ is relevant to our decision making. If the example above, if $\rho = 1$ then almost everyone benefits slightly from the treatment. However if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.

Can these issues be resolved? If so, how?

- note that we can bound this counterfactual distribution based on the variance of the observed interventional distributions. If the variance of XXX is small then this may not be an issue.

- does it actually make sense to make a decision on the basis of this counterfactual that is a function of something we could never observe. I

- part of the problem may be due to the deterministic way we have phrased individual treatment effects.

Can these issues be resolved by considering personalised causal effects as random variables and avoiding counterfactuals?

A range of similar issues can arise in counterfactuals \ref{XXXX}. 
- What assumptions are required for the kind of counterfactual analysis like X would have been higher had B ...?
- briefly discuss the different flavours of counterfactual questions here.

One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation. Rather than taking about the distribution of Y given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $X$. This is certainly convenient, if rather impressive.

@@

Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables. The noise terms are required to be mutually independent. If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable. Structural equation models can be represented visually as a network. Each variable is a node and arrows are drawn from causes to their effects. For example \ref{exm:ranomized_experiment} the SEM is:

\begin{figure}[h]
\caption{SEM for examples \ref{exm:ranomized_experiment}}
\label{fig:causal_network_example2}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = x$};
\node[node, right=of 2](3){$Y = f(X,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\path[]
	(2) edge (3)
	(4) edge (3);
\end{tikzpicture}
\end{figure}

This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$. This is justifiable on the grounds that $X$ is random. The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise). Note that the causal graph in figure \ref{fig:causal_network_example2} is identical to that the bayesian network for the same problem, figure \ref{fig:causal_network_example}. The latent variable $\epsilon_y$ was not explicitly drawn in figure \ref{fig:causal_network_example} as the noise can be captured within the probabilistic node, however adding it would not change the model.

We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'? Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$. The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}. However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$. The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomised experiment. In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.

For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation. However a SEM is, by definition, more than a set of simultaneous equations. By declaring it to be structural we are saying that it represents assumptions about the relationships between variables. When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.

Limitations \& Criticisms 
- can express some non-measurable things
	- example of that

- confusion may occur as to whether interpretation is causal

- can they handle cycles?

@@

Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation). This makes it straightforward to combine key results and algorithms developed within any of these frameworks. For example, draw a graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate. Then use propensity scores \ref{} estimate the effect. If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.

If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables. In other words, the network is a causal Bayesian network. All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.  Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant. This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}). Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks. We can also connect this to counterfactuals via:

@@

We will now demonstrate how we can use our new notation and frameworks for causal inference to resolve a fascinating paradox.  \todo{ insert some history of Simpson's paradox here.}

\cite{simpson1951interpretation}

Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease. She keeps track of which medication her patients choose and whether or not the treatment is successful. She obtains the results in table \ref{tab:simpson1}.

>>>>

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib}
\usepackage{wrapfig}
\usepackage{centernot}

@@

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\cf}[2]{{#1}^{#2}}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

@@

\newcommand{\ci}{\mathrel{\perp\mspace{-10mu}\perp}}
\newcommand{\nci}{\centernot{\ci}}

@@

\begin{document}
\todo{figure out theoremstle is undefined}
%\def\ci{\perp\!\!\!\perp} % from Wikipedia

@@

\tableofcontents

@@

\section{What is causality and why do we care? }

@@

\subsection{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (REFERENCES)
\item what is explanation?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}

\subsection{Identifying when we have a causal problem}

@@

\section{Approaches to causality}

@@

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds. We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectivness of early childhood education including the landmark randomised trial, the Perry Preschool project \citep{weikart1970longitudinal}}
\end{example}

@@

In example \ref{exm:adjusting} we are interested $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing pre-school $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonalble to assume that attending could pre-school effect the likelyhood of graduating from high school and that your parents socio-economic status effects both pre-school attendance and high-school graduation. If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal bayesian network in figure \ref{exm:adjusting}. In this case, the interventional distribution is not equivelent to the observational one. If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool. To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group,

\eqn{
\label{eqn:backdoor_example}
\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}
}

@@

In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would happen if they got the placebo, $\cf{y_{i}}{0}$. The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated. The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment. This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.

Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.  Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution we would see of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate, (from an experimental or observational study);

\eq{
\P{Y|X=1}& \text{, the probability that those who took the treatment will recover}\\
\P{Y|X=0}& \text {, the probability that those who were not treated will recover}
}

Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome. For those who did not take the treatment, the observed outcome is the sames as the outcome \emph{had} they not taken the treatment. Equivalently stated:

\eq{
\P{Y^{0}|X=0}&= \P{Y|X=0}\\
\P{Y^{1}|X=1}&=\P{Y|X=1}
}

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignore-ability assumptions \cite{Rosenbaum1983}. They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated highschool had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$. Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$. A more general form of the ignore-ability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

\vspace*{.3cm}
\begin{theorem}[Ignore-ability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,

\eqn{
\label{eqn:counterfactual1}
\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
\label{eqn:counterfactual2}
\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
}
\end{theorem}

If we assume that within each socio-economic status level, attendence at pre-school is independent to the likelyhood of graduating high-school had you attended then we can compute the average rate of high-school graduation given a universal pre-school program from equation \ref{eqn:counterfactual1}. Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.

Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \cite{Rubin1978}. This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual. As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school. The later school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those did and who now share the classroom with them. SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.

There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised. This makes it quite easy to (accidentally) make statements with counterfactuals that can not be tested with empirical data. Consider the following example based on \citet{Dawid2000}. Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$. Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).

% vector with mean mu_t,mu_c and a covariance matrix
\eqn{
P(\cf{Y}{1},\cf{Y}{0}) \sim N( 
         \begin{bmatrix}
           \mu_{1} \\
           \mu_2 \\
         \end{bmatrix},
         \begin{bmatrix}
           \sigma^2 & \rho \sigma^2 \\
           \rho \sigma^2 & \sigma^2 \\
         \end{bmatrix}
         )
}

Their difference is also normal. Let $\tau = \cf{Y}{1} - \cf{Y}{0}$,

\eqn{
P(\tau) = N(\mu_1 - \mu_0,2\sigma^{2}(1-\rho))
}

\begin{figure}[h]
\captionsetup[subfigure]{position=b}
\centering
\caption{The distribution of individual treatment effects is not identifiable, even from a randomized controlled trial.}
\label{fig:metaphysical_distribution_difference}
\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$. The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated. \label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
\hspace{0.02\textwidth}
\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
\end{figure}

From a (large) randomized controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}. These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated retrospectively. However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}. The key problem is that we can never observe the joint distribution over $\cf{Y}{1}$ and $\cf{Y}{0}$. As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data. \citet{Dawid} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects. He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.

This result is unintuative. It seems on the face of it that the distribution of individual causal effects is relevant to our decision making. If $\rho = 1$ then almost everyone benefits slightly from the treatment whilst if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm. This confusion can be resolved by thinking about personalised rather than individual causal effects. It is entirely possible that potentialy observable characteristics (such as gender, age, genetics, etc) effect how people will respond to the treatment. We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group. The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much we stand to gain from further personalisation. The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).

One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation. Rather than taking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$. This is certainly convenient, if rather imprecise. However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrents careful attention. We should always be clear what assumptions we are making and have in mind if it is possible (at least in theory) to validate those assumptions.

@@

Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables. The noise terms are required to be mutually independent. If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable. Structural equation models can be represented visually as a network. Each variable is a node and arrows are drawn from causes to their effects. Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.

\begin{figure}[h]
\caption{SEM for example \ref{exm:ranomized_experiment}}
\label{fig:sem_randomized_treatment}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(\epsilon_x)$};
\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2);
\end{tikzpicture}
\end{figure}

This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$. This is justifiable on the grounds that $X$ is random. The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise). Note that the causal graph in figure \ref{fig:causal_network_example2} is identical to that the bayesian network for the same problem, figure \ref{fig:causal_network_example}. The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as are captured by the probabilistic nature of the nodes in a Bayesian network, however adding them would not change the model.

Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$. The function $f_y$ and distribution over $\epsilon_y$ does not change. This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convinience},

\eqn {
\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
}

The observational distribution of $Y$ given $X$ is,

\eqn{
\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
}

The interventional distribution is the same as the observational one. The same arguament applies to the intervention $do(X=0)$ and so the causal effect is just the difference in observed outcomes as we found via the causal Bayesian network and counterfactual approaches.

The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}. Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.

\eqn{
\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
\label{eqn:sem:adjusting}
&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
}
Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivelent to the observational distribution, given by;

\eqn{
\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
}

\begin{figure}[h]
\caption{SEM for example \ref{exm:adjusting}}
\label{fig:sem:preschool}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(Z,\epsilon_x)$};
\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
\node[lt, above=of 6](7){$\epsilon_z$};
\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2)
	(7) edge (6)
	(6) edge (2) edge (3);
\end{tikzpicture}
\end{figure}

Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$. A structural equation model with $N$ variables looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation. However a SEM is, by definition, more than a set of simultaneous equations. By declaring it to be structural we are saying that it represents assumptions about the relationships between variables. When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other. The similarity between the notation used to describe and analyse structural equation models and simultaneuos equations, combined with a reluncatnce to make explicit statements about causality has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.

@@

Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation). If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables. In other words, the network is a causal Bayesian network. All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.  Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant. This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}). Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks. We can also connect this to counterfactuals via:

@@

We will now demonstrate how we can apply our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}. The following example is adapted from \citet{Pearl2000}. Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease. She keeps track of which medication her patients choose and whether or not the treatment is successful. She obtains the results in table \ref{tab:simpson1}.
